\documentclass[12pt]{article}
\usepackage{amsmath} % For \eqref
\usepackage{abstract}  
\usepackage[utf8]{inputenc}
\usepackage[letterpaper, margin=1in]{geometry} 
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{tocloft} %Customizers things in contents, list of figures
\usepackage{titlesec}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{makecell}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{float}
\usepackage{afterpage}

\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=teal,
    pdftitle={Physics MSc Thesis},
    pdfpagemode=FullScreen,
    }
% Define a breakable algorithm environment
\makeatletter
\newenvironment{breakablealgorithm}
  {\begin{center}
   \refstepcounter{algorithm}
   \hrule height.8pt depth0pt \kern2pt
   \renewcommand{\caption}[2][\relax]{%
    {\raggedright\textbf{\ALG@name~\thealgorithm} ##2\par}%
    \ifx\relax##1\relax 
      \addcontentsline{loa}{algorithm}{\protect\numberline{\thealgorithm}##2}%
    \else
      \addcontentsline{loa}{algorithm}{\protect\numberline{\thealgorithm}##1}%
    \fi
    \kern2pt\hrule\kern2pt
   }
  }
  {\kern2pt\hrule\relax}
\makeatother


\titleformat{\section}[block]{\Huge\bfseries}{\thesection}{1em}{}
\titlespacing*{\section}{0pt}{1.5em}{5em} 
\titleformat{\subsection}{\Large\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\large\bfseries}{\thesubsubsection}{1em}{}
\renewcommand{\cfttoctitlefont}{\Huge\bfseries} % Adjust font size and style of "Contents"
\renewcommand{\cftloftitlefont}{\Huge\bfseries} % Adjust font size and style of "LOF"
\renewcommand{\cftlottitlefont}{\Huge\bfseries} % Adjust font size and style of "LOT"
\renewcommand{\refname}{\Huge Bibliography\vspace{1.5cm}}

\graphicspath{ {./images/} }
\onehalfspacing
\begin{document}
\setlength{\cftaftertoctitleskip}{2cm} % Customize
\setlength{\cftafterloftitleskip}{2cm} % Customize
\setlength{\cftafterlottitleskip}{2cm} % Customize
% Title Page Block
\begin{titlepage}
    \centering % Center-aligns everything inside the title page
    {\Huge Mountain Waves in the Sky: Overshooting Top Dynamics above Supercell Thunderstorms \par} % Large title
    \vspace{2cm}
    {\huge Author: Parveer Banwait \par}
    \vspace{0.25cm}
    {\large Student Number: 1007016499 \par}
    \vspace{1cm}
    {\Large Supervised by: Professor Morgan O'Neill \par}
    \vfill % Move proceeding text to bottom of the page
    {\Large August 2026 \par}
    \vspace{0.25cm}
    {\Large MSc Thesis \par}
    \vspace{0.25cm}
    {\Large Final Report \par}
\end{titlepage}
\pagenumbering{roman}
\vspace*{\fill}
\begin{abstract}
\begin{spacing}{2.0}
Abstract
\end{spacing}
\end{abstract}
\vspace*{\fill}
\newpage
\renewcommand{\abstractname}{Acknowledgements}  % Change heading name
\vspace*{\fill}
\begin{abstract}
I would like to express my gratitude to my supervisor Morgan O'Neill for ....
\end{abstract}
\vspace*{\fill}

\newpage

\tableofcontents

\newpage

\listoffigures

\newpage

\listoftables

\newpage

\pagenumbering{arabic}
\section{Introduction}

\subsection{AACP Overview}
Stratospheric water vapour (SWV) is an important climate feedback, with the ability to potentially impact surface warming \cite{dessler_stratospheric_2013}, hamper ozone recovery \cite{shindell_climate_2001} and decrease the intensity of the stratospheric circulation \cite{veenus_observational_2025}, influencing the global ozone distribution. Thus, it is critical to understand how the SWV budget in order to quantify these feedbacks. Unfortunately, global climate models currently struggle with this, with an average 0.5 ppmv underestimation of ppmv at 70hPa alongside a large model spread \cite{keeble_evaluating_2021}. An important secondary source of SWV is mid-latitude convection from supercell thunderstorms. These storms produce overshooting tops (OTs), deep convection which overshoots the level of neutral buoyancy and penetrates up into the stratosphere. With strong storm-relative prevailing winds in the stratosphere large amounts of water vapour within this OT can be injected irreversibly into the stratosphere in the form of a turbulent hydraulic jump \cite{oneill_hydraulic_2021}.

The SWV feedback on surface temperature has been a source of contention in recent times. In 2001, \cite{shindell_climate_2001} found that the combined water vapour - ozone feedback due to changes from 1959 to 1999 led to a net 0.07 Kelvin surface warming per decade. This was despite the ozone feedback on its own having a net cooling effect. In 2013, \cite{dessler_stratospheric_2013} demonstrated that increases in SWV due to increases in tropospheric temperature causes a $0.3 W/m^2$ increase in surface flux per Kelvin of surface warming. A third of this was found to be due to water vapour entering the stratosphere through the tropical tropopause layer, with the remaining two thirds being due to water vapour entering through extra-tropics. However, recent studies challenge this claim. For example, using a multi-model analysis, He at al. (2025) \cite{he_muted_2025} claimed that after both stratospheric cooling and tropospheric warming feedbacks associated with SWV changes are included, the total radiative forcing due to SWV is negligible. The uncertainty around the radiative feedback of SWV further increases the importance of accurately calculating its budget, and finding how it will change under climate change, for future studies.

Water vapour can travel up into the stratosphere in many different ways. The main method of transport is through the tropical cold point tropopause \cite{keeble_evaluating_2021}, although transport from volcanic eruptions \cite{millan_hunga_2022} and convective overshooting are important secondary sources. Focusing on convective overshooting, \cite{dauhut_mechanisms_2018} uses a model to show that tropical convective storms, such as the "Hector the Convector" can inject over 2700 tons of water vapour into the stratosphere over the course of an hour, with this injection being poorly resolved by convective parameterizations. In the midlatitudes, an aircraft flown 20 hours after a supercell thunderstorm showed that large 1-2 ppmv enhancements in water vapour were found well above the tropopause (19.25 km up!) \cite{homeyer_extreme_2023}. Additionally, model simulations have shown that water vapour injected into the stratospheric overworld (above the 380K potential temperature isoline) by OTs tends to stay even after the supercell system which triggered them is gone \cite{wang_convectively_2023}. This injection is not well resolved by current generation satellites such as Aura's Microwave Limb Sounder.

In about 10\% of cases, storms produce an above anvil cirrus plume (AACP), with AACP's being formed by 75\% of supercells. Despite this, AACPs produce 57\% of all severe weather reports and 73\% of significant severe weather reports \cite{bedka_above-anvil_2018}. AACP's are characterized by a unique texture distinct from the storm larger anvil several kilometers underneath, as well as as a high IR brightness temperature (but not always)! AACPs precede severe weather reports by 31 minutes on average, and this one feature predates  U.S National Weather Service warnings 25\% of the time. It has also been shown that AACP storms inject more water into the stratospheric overworld \cite{oneill_hydraulic_2021} \cite{gordon_sensitivities_2022} \cite{homeyer_development_2017}, although whether they inject more into the stratospheric middleworld is disputed. High storm-relative winds in the lower stratosphere and increasing low-level instability have been found to be environmental conditions which favour AACP formation \cite{homeyer_development_2017}.

The mechanism behind AACP formation is the breaking of gravity waves forced by the OT \cite{oneill_hydraulic_2021} \cite{dauhut_mechanisms_2018} \cite{wang_moisture_2003} \cite{hassim_model_2010} \cite{qu_simulation_2020}. The OT acts similarly to a mountain at the surface, blocking the incoming stratiform flow, and forcing the winds up and over the OT. The flow then accelerates over this "mountain", continuing to accelerate as it travels over the leeside of the OT. These waves break, and the accelerated flow reconnects to the ambient downstream flow in the form of a turbulent hydraulic jump \cite{oneill_hydraulic_2021}. This hydraulic jump takes the high levels of water perturbation from the OT (which originated in the upper troposphere \cite{hassim_model_2010}) and mixes it into the stratosphere (see O'Neill et al. (2021) \cite{oneill_hydraulic_2021} figure 2B). The cloud ice "jumped" from the OT is what forms the AACP. In subsaturated environments, this cloud ice will then sublimate, irreversibly adding water vapour into the stratosphere. However, in supersaturated environments the vapour will actually deposit on the ice crystals causing the stratosphere to lose moisture when the ice precipitates out \cite{hassim_model_2010}. When the storm ends, a lot of this water vapour which hasn't been mixed in this fashion will fall back into the troposphere with the OT \cite{wang_convectively_2023}. In their 1km grid spacing simulation, Qu et al. (2020) \cite{qu_simulation_2020} found that wave breaking accounted for 59\% of the vertical vapour injection via vertical transport while the mean updraft accounted for 39\%. However, this ratio is expected to skew further towards wave breaking as the simulation resolution is increased as more wave breaking is resolved.

\subsection{Mountain Waves}











\newpage
\bibliographystyle{unsrt} % You can use other styles like 'unsrt', 'alpha', etc.
\bibliography{references}

\end{document}